---
title: "Additional Sensitivity Analyses"
author: "Denis Haine"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Additional Sensitivity Analyses}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Other functions could be useful and are not part of the book
["Applying Quantitative Bias Analysis to Epidemiologic
Data"](https://www.springer.com/us/book/9780387879604).
These functions are:

* `confounders.array`: Bias due to unmeasured confounders based on confounding imbalance among exposed and unexposed
* `confounders.ext`: Unmeasured confounders based on external measurement
* `confounders.evalue`: E-value due to unmeasured confounder

## Array Approach for Unmeasured Confounders

`episensr` also offers the option to check the effect of residual confounding
based on the array approach ([Schneeweiss,
2006](https://onlinelibrary.wiley.com/doi/abs/10.1002/pds.1200)).

Even after controlling for confounding (in the design or analysis of a study),
some residual confounding  can still be present, because:

1. some confounding factors were not taken into account (not looked at, not
    adjusted for, or no data collected about them),
2. you only loosely took into account that confounding (still presence of
    differences between groups after control by matching for example, or you
    lost precision on a confounding variable e.g. by categorizing a continuous
    variable),
3. subjects were not correctly classified to confounding variables.

### Simple Sensitivity Analyses in the Absence of External Information: Array Approach

The *array approach* by Schneeweiss provides a corrected RR based on the
apparent, confounded, RR, the prevalence of the confounding factor among the
exposed and unexposed, and the association between the confounder and the outcome.

Schneeweiss provides the (made-up) example of the association between
TNF\(_\alpha\) blocking agents and Non-Hodgkin (NH) Lymphoma in rheumatoid
arthritis (RA) patients.
Assuming an observed relative risk of 2, that 50% of RA patients have a more
progressive immunologic disease, and that more progressive disease is more
likely to lead to NH lymphoma:

```{r conf_array}
library(episensr)
confounders.array(crude.risk = 2,
                  type = "binary",
                  bias_parms = c(2.5, 0.1, 0.5))
```

The function takes values of the observed risk (2.0); if the covariates are
binary, continuous, or on a risk difference scale; the association between
confounder and outcome (let's say here 2.5); the prevalence of the confounder
among the exposed (let's say 0.1); and the prevalence of the confounder among
the unexposed.
The adjusted RR is 3.04.

However, it is more interesting if we could keep some factors constant in our
analysis (the crude risk of 2.0, the prevalence among the unexposed, 0.5), and
vary the imbalance of the hypothetical unobserved confounder (association
confounder-outcome from 1 to 5.5; prevalence among exposed, from 0 to 1):

```{r conf_array_apply,warning=FALSE,message=FALSE,fig.width=6,fig.height=4}
dat <- expand.grid(RR_CD = seq(1, 5.5, 0.5), P_C1 = seq(0, 1, 0.1))
dat$RR_adj <- apply(dat, 1,
                    function(x) confounders.array(2,
                                                  type = "binary",
                                                  bias_parms = c(x[1],
                                                                 x[2],
                                                                 0.5),
                                                  print = FALSE)[5])
```

Then adding fun with 3-d plots!

```{r array_3d,warning=FALSE,message=FALSE,fig.width=8,fig.height=8}
library(lattice)
library(grid)
wireframe(RR_adj ~ RR_CD * P_C1,
          data = dat,
          xlab = "RR\nconfounder-outcome",
          ylab = "Prevalence confounder\namong exposed",
          zlab = "Adjusted\nRR",
          drape = TRUE, colorkey = TRUE,
          scales = list(arrows = FALSE, cex = .5, tick.number = 10,
                        z = list(arrows = FALSE), distance = c(1.5, 1.5, 1.5)),
          zlim = 0:4,
          light.source = c(10, 0, 10),
          col.regions = rainbow(100, s = 1, v = 1,
                                start = 0, end = max(1, 100 - 1) / 100,
                                alpha = .8),
          screen = list(z = -60, x = -60),
          panel = function(...)
          {
              panel.wireframe(...)
              grid.text("Observed RR = 1\nPrevalence confounder\namong unexposed = 0.5",
                        0.125, 0.175, default.units = "native")
          })
```

### Using Additional Information: External Adjustment

You can also use additional information not available in the main study to
adjust for the effect of confounding.
Sometimes you can get information from a representative sample to quantify the
imbalance risk factors that are measured among the exposed group; and from the
literature (RCT, observational studies) for the association between the
confounder and the outcome.

We can use the example of Schneeweiss regarding the association between
selective COX-2 inhibitor use and the incidence of MI.

Say we assume an exposure-disease association of 1 (null hypothesis).
The more the truth is away from the null, the more bias in our estimate.
However the less relevant unmeasured confounders become.

Let's try:

```{r conf_ext}
confounders.ext(RR = 1, bias_parms = c(0.1, 1.6, 0.1, 0.51))
```

Same here, the interest lies in using the function over a range of values, while
keeping some other constant.
And we can extend to several confounders:

```{r conf_ext_apply,warning=FALSE,message=FALSE,fig.width=8,fig.height=6}
dat <- expand.grid(RR_CD = seq(0.1, 1, 0.1))
dat$nsaid <- apply(dat, 1,
                   function(x) confounders.ext(1,
                                               bias_parms = c(x[1], 0.9, 0.1, 0.4),
                                               print = FALSE)[7])
dat$non_user <- apply(dat, 1,
                      function(x) confounders.ext(1,
                                                  bias_parms = c(x[1], 1.03, 0.09,
                                                                 0.12),
                                                  print = FALSE)[7])
dat$naproxen <- apply(dat, 1,
                      function(x) confounders.ext(1,
                                                  bias_parms = c(x[1], 1.15, 0.09,
                                                                 0.79),
                                               print = FALSE)[7])
dat$rof_napro <- apply(dat, 1,
                       function(x) confounders.ext(1,
                                               bias_parms = c(x[1], 1.6, 0.1, 0.51),
                                               print = FALSE)[7])
library(tidyverse)
dat2 <- dat %>% gather(nsaid, non_user, naproxen, rof_napro,
                       key = "COX2", value = "bias_perc")

ggplot(dat2, aes(x = RR_CD, y = bias_perc, group = COX2, colour = COX2)) +
    geom_line() +
    scale_colour_brewer(palette = "Dark2",
                        labels = c("COX-2 vs. naproxen", "COX-2 vs. non-users",
                                   "COX-2 vs. non-selective NSAIDs",
                                   "Rofecoxib vs. naproxen")) +
    geom_vline(xintercept = 0.7) +
    xlab("Association Confounder-Disease (RR)") +
    ylab("Bias of Exposure-Disease Association (RR) in %") +
    ggtitle(expression(paste("Bias as a function of misspecification of the ", RR[CD], " from the literature"))) +
    theme(legend.position = c(.85, .3),
          legend.title = element_blank()) +
    annotate("text", label = expression(paste("Literature estimate ", RR[CD], " = 0.7")), x = 0.6, y = -3.75)
```

## Unmeasured Confounding and E-value

E-value, introduced by [VanderWeele et
al.](https://pubmed.ncbi.nlm.nih.gov/28693043/), is the minimum strength
of association on the risk ratio scale that an unmeasured confounder would need
to have with both the exposure and the outcome, conditional on the measured
covariates, to fully explain away a specific exposure-outcome association.

For example, a study by [Victoria et
al.](https://www.thelancet.com/journals/lancet/article/PIIS0140-6736(87)90902-0/fulltext)
found an association between formula-fed babies compared to breast-fed and
respiratory death, RR = 3.9. What would be the strength of association for an
unmeasured confounder needed to explain away this association? The E-value
express the magnitude of the confounder associations that can produce
confounding bias equal to the observed exposure-outcome association. This E-value is:

```{r e-value-1}
confounders.evalue(est = 3.9, lower_ci = 1.8, upper_ci = 8.7, type = "RR")
```

The E-value is 7.26, i.e. an unmeasured confounder must have a risk ratio of at
least 7.26 with both the exposure and the outcome, and above and beyond the
measured confounders, to explain away the observed risk ratio of 3.9. A weaker
unmeasured confounder could not explain this RR. E-value for the CI closest to
the null (i.e. 1) can also be reported.

Similarly, the minimum magnitude of both confounder associations to move away
the observed association can be assessed for a non-null hypothesis. For example, the
effect of anti-depressant use during pregnancy on infant cardiac defects was
1.06 [0.93--1.22] ([Huybrechts et
al.](https://www.nejm.org/doi/full/10.1056/NEJMoa1312828)). The confidence
interval includes 1 so no further unmeasured confounding is needed to shift the
confidence interval to include 1, i.e. the E-value = 1. But what would be the
strength of a confounder needed to shift RR to 1.2?

```{r e-value-2}
confounders.evalue(est = 1.06, lower_ci = 0.93, upper_ci = 1.22,
                   type = "RR", true_est = 1.2)
```

We need an unmeasured confounder with a RR of 1.5 each with the outcome, cardiac
defect, and the exposure, anti-depressant use, to move the risk ratio to 1.2. A
weaker confounder could not shift the observed RR.

E-value can also be computed when the effect measure is:

* OR or HR for rare outcomes (i.e. < 15%), rate ratio for count and continuous
  outcomes: `RR`,
* OR for common outcomes (i.e. > 15%): `ORc`,
* HR for common outcomes (i.e. > 15%): `HRc`,
* Difference in continuous outcomes, with standardized effect size *d* (mean of
  the outcome variable divided by the SD of the outcome) and an SE for this
  standardized effect size *s_d*, either with RR approximation or OR
  approximation (meta-analysis): `diff_RR` and `diff_OR`.
