---
title: "Quantitative Bias Analysis for Epidemiologic Data"
author: "Denis Haine"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Quantitative Bias Analysis for Epidemiologic Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Quantitative bias analysis allows to estimate nonrandom errors in epidemiologic
studies, assessing the magnitude and direction of biases, and quantifying their
uncertainties.
Every study has some random error due to its limited sample size, and is
susceptible to systematic errors as well, from selection bias to the presence of
(un)known confounders or information bias (measurement error, including
misclassification).
Bias analysis methods were compiled by Lash et al. in their
book
["Applying Quantitative Bias Analysis to Epidemiologic Data"](https://www.springer.com/us/book/9780387879604).
This package implements the various bias analyses from that book, which are also
[available](https://sites.google.com/site/biasanalysis/) (for some) as a
spreadsheet or a SAS macro.
This vignette provides some examples on how to use the package.

## Selection Bias

We will use a case-control study by
[Stang et al.](http://www.ncbi.nlm.nih.gov/pubmed/16523014) on the relation
between mobile phone use and uveal melanoma.
The observed odds ratio for the association between regular mobile phone use vs.
no mobile phone use with uveal melanoma incidence is 0.71 [95% CI 0.51-0.97].
But there was a substantial difference in participation rates between cases and
controls (94% vs 55%, respectively) and so selection bias could have an impact
on the association estimate.
The 2X2 table for this study is the following:

|          | Regular use | No use |
|---------:|:-----------:|:------:|
| Cases    | 136         | 107    |
| Controls | 297         | 165    |

We use the function `selection` as shown below.

```{r selection}
library(episensr)

stang <- selection(matrix(c(136, 107, 297, 165),
                          dimnames = list(c("UM+", "UM-"), c("Mobile+", "Mobile-")),
                          nrow = 2, byrow = TRUE),
                   bias_parms = c(.94, .85, .64, .25))
stang
```

The various `episensr` functions return an object which is a list containing the
input and output variables.
You can check it out with `str()`.

The 2X2 table is provided as a matrix and selection probabilities given with the
argument `bias_parms`, a vector with the 4 probabilities (guided by the
participation rates in cases and controls) in the following order: among cases
exposed, among cases unexposed, among noncases exposed, and among noncases
unexposed.
The output shows the observed 2X2 table and the observed odds ratio (and relative
risk), followed by the corrected ones.

## Uncontrolled Confounders

We will use data from a cross-sectional study by
[Tyndall et al.](http://www.ncbi.nlm.nih.gov/pubmed/8879763) on the association
between male circumcision and the risk of acquiring HIV, which might be
confounded by religion.
The code to account for unmeasured or unknown confounders is the following,
where the 2X2 table is given as a matrix.
We choose a risk ratio implementation, provide a vector defining the risk ratio
associating the confounder with the disease, and the prevalence of the
confounder, religion, among the exposed and the unexposed.

```{r confounders}
confounders(matrix(c(105, 85, 527, 93),
                   dimnames = list(c("HIV+", "HIV-"), c("Circ+", "Circ-")),
                   nrow = 2, byrow = TRUE),
            type = "RR",
            bias_parms = c(.63, .8, .05))
```

The output gives the crude 2X2 table, the crude relative risk and confounder specific measures of
association between exposure and outcome, and the relationship adjusted for the
unknown confounder, using a standardized morbidity ratio (SMR) or a
Mantel-Haenszel (MH) estimate of the risk ratio.

## Probabilistic Sensitivity Analysis for Exposure Misclassification

We use a study on the effect of smoking during pregnancy on breast cancer risk
by [Fink & Lash](http://www.ncbi.nlm.nih.gov/pubmed/12946045), where we assume
nondifferential misclassification of the exposure, smoking, with probability
density functions for sensitivities (Se) and specificities (Sp) among cases and
noncases equal to uniform distributions with a minimum of 0.7 and a maximum of
0.95 for sensitivities (0.9 and 0.99 respectively for specificities).
We choose to correct for exposure misclassification with the argument `type =
exposure`.
We ask for 10000 replications (default is 1000).
The Se and Sp for cases (`seca`, `spca`) are given as a list with its first
element referring to the type of distribution (choice between constant, uniform,
triangular, trapezoidal, logit-logistic, and logit-normal) and the second
element giving the distribution parameters (min and max for uniform
distribution).
By avoiding to provide information on the noncases (`seexp`, `spexp`), we are
referring to a nondifferential misclassification.

```{r probsens}
set.seed(123)
smoke.nd <- probsens(matrix(c(215, 1449, 668, 4296),
                            dimnames = list(c("BC+", "BC-"), c("Smoke+", "Smoke-")),
                            nrow = 2, byrow = TRUE),
                     type = "exposure",
                     reps = 10000,
                     seca.parms = list("uniform", c(.7, .95)),
                     spca.parms = list("uniform", c(.9, .99)))
smoke.nd
```

The output gives the 2X2 table, the observed measures of association, and the
corrected measures of association.

We saved the `probsens` analysis in a new object `smoke.nd`.
We can see its elements with the function `str()`:

```{r str}
str(smoke.nd)
```

`smoke.nd` is a list of 4 elements where different information on the analysis
done are saved.
We have `smoke.nd$obs.data` where we have the observed 2X2 table,
`smoke.nd$obs.measures` (the observed measures of association),
`smoke.nd$adj.measures` (the adjusted measures of association), and
`smoke.nd$sim.df`, a data frame with the simulated variables from each
replication, like the Se and Sp, the 4 cells of the adjusted 2X2 table, and the
adjusted measures.
We can plot the Se prior distribution (and not forgetting to discard the draws that
led to negative adjustments).

```{r plot, fig.cap = "Sensibility prior distribution."}
hist(smoke.nd$sim.df[!is.na(smoke.nd$sim.df$corr.RR), ]$seca,
     breaks = seq(0.65, 1, 0.01),
     col = "lightgreen",
     main = NULL,
     xlab = "Sensitivity for Cases")
```

## Probabilistic Sensitivity Analysis for Uncontrolled Confounding

Let's illustrate this function with this example from [Modern Epidemiology by
Rothman, Greenland &
Lash](https://shop.lww.com/Modern-Epidemiology/p/9781451190052), where the
association between occupational resins exposure and lung cancer mortality is
studied, together with the presence of an unmeasured potential confounder,
smoking ([Greenland et al., 1994](https://www.ncbi.nlm.nih.gov/pubmed/7927843)).

|          | Resins exposed | Resins unexposed |
|---------:|:--------------:|:----------------:|
|    Cases | 45             | 94               |
| Controls | 257            | 945              |

Prior probability distributions are given to each bias parameters.
Prevalences of smoking in those exposed to resins, and of smoking in those
unexposed to resins receive prior distributions that are uniform between 0.40
and 0.70. 
Prior distribution for the odds ratio associating smoking with lung cancer
mortality is log-normal with 95% limits of 5 and 15.
The mean of this distribution is [ln(15) + ln(5)] / 2 = 2.159, and its standard
deviation is [ln(15) - ln(5)] / 3.92 = 0.28.

```{r probsens-conf}
set.seed(123)
probsens.conf(matrix(c(45, 94, 257, 945),
                     dimnames = list(c("Cases+", "Cases-"), c("Res+", "Res-")),
                     nrow = 2, byrow = TRUE),
              reps = 50000,
              prev.exp = list("uniform", c(.4, .7)),
              prev.nexp = list("uniform", c(.4, .7)),
              risk = list("log-normal", c(2.159, .28)))
```

The median adjusted OR is 1.76 [1.25,2.49].

## Bootstrapping

Selection and misclassification bias models can be bootstrapped in order to get
confidence interval on the adjusted association parameter.
We can use the ICU dataset from Hosmer and
Lemeshow
[Applied Logistic Regression](http://ca.wiley.com/WileyCDA/WileyTitle/productCd-0470582472.html) textbook
as an example.
The replicates that give negative cells in the adjusted 2X2 table are silently
ignored and the number of effective bootstrap replicates is provided in the
output.

```{r, boot}
library(aplore3)  # to get ICU data
data(icu)

## First run the model
misclass_eval <- misclassification(icu$sta, icu$inf,
                                   type = "exposure",
                                   bias_parms = c(.75, .85, .9, .95))
misclass_eval

## Then bootstrap it
set.seed(123)
misclass_boot <- boot.bias(misclass_eval, R = 10000)
misclass_boot
```

Bootstrap replicates can also be plotted, with the confidence interval shown as
dotted lines.

```{r, boot_fig, fig.cap = "Bootstrap replicates and confidence interval.", warning=F}
plot(misclass_boot, "rr")
```

