---
title: "Quantitative Bias Analysis for Epidemiologic Data"
author: "Denis Haine"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Quantitative Bias Analysis for Epidemiologic Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Quantitative bias analysis allows to estimate nonrandom errors in epidemiologic
studies, assessing the magnitude and direction of biases, and quantifying their
uncertainties.
Every study has some random error due to its limited sample size, and is
susceptible to systematic errors as well, from selection bias to the presence of
(un)known confounders or information bias (measurement error, including
misclassification).
Bias analysis methods were compiled by Lash et al. in their
book
["Applying Quantitative Bias Analysis to Epidemiologic Data"](https://link.springer.com/book/10.1007/978-3-030-82673-4).
This package implements the various bias analyses from that book, which are also
[available](https://sites.google.com/site/biasanalysis/) (for some) as a
spreadsheet or a SAS macro, as well as some additional approaches.
This vignette provides some examples on how to use the package.

Functions available in `episensr` are:

* `selection`: Selection bias
* `mbias`: Selection bias caused by M bias
* `confounders`: Unmeasured or unknown confounders
* `confounders.poly`: Polytomous confounders
* `confounders.emm`: Unmeasured or unknown confounders in the presence of effect modification
* `confounders.limit`: Bounding the bias limits of unmeasured confounding
* `confounders.array`: Bias due to unmeasured confounders based on confounding imbalance among exposed and unexposed
* `confounders.ext`: Unmeasured confounders based on external measurement
* `confounders.evalue`: E-value due to unmeasured confounder
* `misclassification`: Disease or exposure misclassification
* `misclassification_cov`: Covariate misclassification
* `bootstrap`: Bootstrap resampling for selection and misclassification bias
* `multidimBias`: Multidimensional bias analysis
* `probsens.sel`: Probabilistic analysis for selection bias
* `probsens.conf`: Probabilistic analysis for unmeasured confounding
* `probsens`: Probabilistic analysis for misclassification
* `probsens.irr.conf`: Probabilistic analysis for unmeasured confounding of person-time data
* `probsens.irr`: Probabilistic analysis for exposure misclassification of person-time data
* `multiple bias`: Multiple bias modeling

## Selection Bias

We will use a case-control study by
[Stang et al.](https://pubmed.ncbi.nlm.nih.gov/16523014/) on the relation
between mobile phone use and uveal melanoma.
The observed odds ratio for the association between regular mobile phone use vs.
no mobile phone use with uveal melanoma incidence is 0.71 [95% CI 0.51-0.97].
But there was a substantial difference in participation rates between cases and
controls (94% vs 55%, respectively) and so selection bias could have an impact
on the association estimate.
The 2X2 table for this study is the following:

|          | Regular use | No use |
|---------:|:-----------:|:------:|
| Cases    | 136         | 107    |
| Controls | 297         | 165    |

We use the function `selection` as shown below.

```{r selection}
library(episensr)

stang <- selection(matrix(c(136, 107, 297, 165),
                          dimnames = list(c("UM+", "UM-"), c("Mobile+", "Mobile-")),
                          nrow = 2, byrow = TRUE),
                   bias_parms = c(.94, .85, .64, .25))
stang
```

The various `episensr` functions return an object which is a list containing the
input and output variables.
You can check it out with `str()`.

The 2X2 table is provided as a matrix and selection probabilities given with the
argument `bias_parms`, a vector with the 4 probabilities (guided by the
participation rates in cases and controls) in the following order: among cases
exposed, among cases unexposed, among noncases exposed, and among noncases
unexposed.
The output shows the observed 2X2 table and the observed odds ratio (and relative
risk), followed by the corrected ones.

## Misclassification Bias

Misclassification bias can be assessed with the function `misclassification`.
Confidence intervals for corrected association due to exposure misclassification
are also directly available, or the estimates can also be bootstrapped (see below).
The confidence intervals from the variance of the corrected odds ratio estimator in the
`misclassification` function are computed as in [Greenland et
al.]( https://doi.org/10.1002/sim.4780070704) and [Chu et
al.](https://doi.org/10.1016/j.annepidem.2006.04.001), when adjusting for exposure misclassification
using sensitivity and specificity.
Using the example in Chu et al. of a case-control study of cigarette smoking and
invasive pneumococcal disease, the unadjusted odds ratio is 4.32, with a 95%
confidence interval of 2.96 to 6.31.
Let's say the sensitivity of self-reported smoking is 94% and specificity is
97%, for both the case and control groups:

```{r misclass}
misclassification(matrix(c(126, 92, 71, 224),
                         dimnames = list(c("Case", "Control"),
                                         c("Smoking +", "Smoking - ")),
                         nrow = 2, byrow = TRUE),
                  type = "exposure",
                  bias_parms = c(0.94, 0.94, 0.97, 0.97))
```

The corrected odds ratio is now 5.02, with a widened 95% confidence interval
(3.28 to 7.69).
Note the bias despite the large sensitivity and specificity.

You can even reproduce the contour plots in Chu et al. paper!

```{r, warning=FALSE,fig.cap='Contour plot of point estimates of corrected odds ratio (OR)',fig.width=6,fig.height=4}
dat <- expand.grid(Se = seq(0.582, 1, 0.02),
                   Sp = seq(0.762, 1, 0.02))

dat$OR_c <- apply(dat, 1,
                  function(x) misclassification(matrix(c(126, 92, 71, 224),
                                                       nrow = 2, byrow = TRUE),
                                                type = "exposure",
                                                bias_parms = c(x[1], x[1], x[2], x[2]))$adj.measures[2, 1])
dat$OR_c <- round(dat$OR_c, 2)

library(ggplot2)
library(directlabels)
p1 <- ggplot(dat, aes(x = Se, y = Sp, z = OR_c)) +
    geom_contour(aes(colour = ..level..), breaks = c(4.32, 6.96, 8.56, 12.79, 23.41, 432.1)) +
    annotate("text", x = 1, y = 1, label = "4.32", size = 3) +
    scale_fill_gradient(limits = range(dat$OR_c), high = 'red', low = 'green') +
    scale_x_continuous(breaks = seq(0.5, 1, .1), expand = c(0,0)) +
    scale_y_continuous(breaks = seq(0.5, 1, .1), expand = c(0,0)) +
    coord_fixed(ylim = c(0.5, 1.025), xlim = c(0.5, 1.025)) +
    scale_colour_gradient(guide = 'none') +
    xlab("Sensitivity") +
    ylab("Specificity") +
    ggtitle("Estimates of Corrected OR")
direct.label(p1, list("far.from.others.borders", "calc.boxes", "enlarge.box",
                      hjust = 1, vjust = -.5, box.color = NA, cex = .6,
                      fill = "transparent", "draw.rects"))
```

```{r, warning=FALSE,fig.cap='Contour plot of 95% lower confidence limit of corrected OR',warning=FALSE,fig.width=6,fig.height=4}
dat$ORc_lci <- apply(dat, 1,
                     function(x) misclassification(matrix(c(126, 92, 71, 224),
                                                          nrow = 2, byrow = TRUE),
                                                   type = "exposure",
                                                   bias_parms = c(x[1], x[1], x[2], x[2]))$adj.measures[2, 2])
dat$ORc_lci <- round(dat$ORc_lci, 2)

p3 <- ggplot(dat, aes(x = Se, y = Sp, z = ORc_lci)) +
    geom_contour(aes(colour = ..level..),
                 breaks = c(4.05, 4.64, 5.68, 7.00, 9.60)) +
    annotate("text", x = 1, y = 1, label = "2.96", size = 3) +
    scale_fill_gradient(limits = range(dat$ORc_lci), high = 'red', low = 'green') +
    scale_x_continuous(breaks = seq(0.5, 1, .1), expand = c(0,0)) +
    scale_y_continuous(breaks = seq(0.5, 1, .1), expand = c(0,0)) +
    coord_fixed(ylim = c(0.5, 1.025), xlim = c(0.5, 1.025)) +
    scale_colour_gradient(guide = 'none') +
    xlab("Sensitivity") +
    ylab("Specificity") +
    ggtitle("95% LCI of Corrected OR")
direct.label(p3, list("far.from.others.borders", "calc.boxes", "enlarge.box",
                           hjust = 1, vjust = -.5, box.color = NA, cex = .6,
                           fill = "transparent", "draw.rects"))
```

The bias analysis for exposure misclassification can also use positive and negative predictive values.
[Bodnar et al.](https://doi.org/10.1111/ppe.12120) evaluated the accuracy of maternal prepregnancy BMI
and gestational weight gain (GWG) data derived from the Pennsylvania state birth certiï¬cates against
information collected from the medical record. To estimate positive and negative predictive values, a
validation study was conducted by randomly sampling women conditional on their reported BMI category
and term/preterm status. BMI was obtained from medical records for these sampled women and used as a
gold standard for BMI category, allowing to determine positive and negative predictive values:
PPV<sub>D1</sub> = 65%, PPV<sub>D0</sub> = 74%, NPV<sub>D1</sub> = 100%, NPV<sub>D0</sub> = 98%. The
analysis is the following.

```{r, misclass_pv}
misclassification(matrix(c(599, 4978, 31175, 391851),
                         dimnames = list(c("Preterm", "Term"), c("Underweight", "Normal weight")),
                         nrow = 2, byrow = TRUE),
                  type = "exposure_pv",
                  bias_parms = c(0.65, 0.74, 1, 0.98))
```

Note that using predictive values relates to the prevalence of the true exposure status, which might
not be the same in every populations.

Covariate misclassification is available via the function `misclassification.cov`.
For example, the paper by [Berry et
al.](https://www.bmj.com/content/330/7495/815/) looked if misclassification of
the confounding variable _in vitro fertilization_ (IVF), a confounder, resulted
wrongly on an association between increase folic acid and having twins.
IVF increases the risk of twins, and women undergoing IVF might be more likely
to take folic acid supplements, i.e. IVF would be a confounder between vitamins
and twins.
Data on IVF were not available and a proxy for it was used, _period of
involuntary childlessness_.
However, it was a poor proxy for IVF, with a sensitivity of 60% and a
specificity of 95%.
These bias parameters were assumed to be nondifferential.
Here's the code with `episensr`:

```{r cov_misclass}
misclassification.cov(array(c(1319, 38054, 5641, 405546, 565, 3583,
                              781, 21958, 754, 34471, 4860, 383588),
                            dimnames = list(c("Twins+", "Twins-"),
                                            c("Folic acid+", "Folic acid-"),
                                            c("Total", "IVF+", "IVF-")),
                            dim = c(2, 2, 3)),
                      bias_parms = c(.6, .6, .95, .95))
```

While the non-adjusted analysis showed that women taking folic acid were 2.44
times more likely to have twins, after correcting for the misclassification of
IVF, risk ratio is now null (= 1.0).

## Uncontrolled Confounders

We will use data from a cross-sectional study by
[Tyndall et al.](https://pubmed.ncbi.nlm.nih.gov/8879763/) on the association
between male circumcision and the risk of acquiring HIV, which might be
confounded by religion.
The code to account for unmeasured or unknown confounders is the following,
where the 2X2 table is given as a matrix.
We choose a risk ratio implementation, provide a vector defining the risk ratio
associating the confounder with the disease, and the prevalence of the
confounder, religion, among the exposed and the unexposed.

```{r confounders}
tyndall <- confounders(matrix(c(105, 85, 527, 93),
                              dimnames = list(c("HIV+", "HIV-"), c("Circ+", "Circ-")),
                              nrow = 2, byrow = TRUE),
                       type = "RR",
                       bias_parms = c(.63, .8, .05))
tyndall
```

The output gives the crude 2X2 table, the crude relative risk and confounder specific measures of
association between exposure and outcome, and the relationship adjusted for the
unknown confounder, using a standardized morbidity ratio (SMR) or a
Mantel-Haenszel (MH) estimate of the risk ratio.

Additional information are also available, like the 2X2 tables by levels of the
confounder.

```{r confounders_add}
## Confounder +
tyndall$cfder.data

## Confounder -
tyndall$nocfder.data
```

As described in this function's help file, the bias correction used is the
"relative risk due to confounding" as proposed by Miettinen (Components of the
crude risk ratio, Am J Epidemiol 1972, 96(2):168-172).
The two examples shown in that paper are the following.

The first one is about drug surveillance data analyzed as a cohort study, with
the frequency of drug-attributed rash in relation to allopurinol exposure among
recipients of ampicillin, with sex treated as a potential confounding factor.

```{r miettinen1}
rash <- matrix(c(15, 94, 52, 1163),
               dimnames = list(c("Rash +", "Rash -"),
                               c("Allopurinol +", "Allopurinol -")),
               nrow = 2, byrow = TRUE)
```

```{r miettinen1-tab, echo=FALSE}
knitr::kable(rash)
```

We can decompose these numbers by confounders:

```{r miettinen1-conf}
## Outcome by confounders
rash_conf <- matrix(c(36, 58, 645, 518),
                    dimnames = list(c("Rash +", "Rash -"),
                                    c("Males", "Females")),
                    nrow = 2, byrow = TRUE)
## By confounders: among males
rash_males <- matrix(c(5, 36, 33, 645),
                     dimnames = list(c("Rash +", "Rash -"),
                                     c("Allopurinol +", "Allopurinol -")),
                     nrow = 2, byrow = TRUE)
## By confounders: among females
rash_females <- matrix(c(10, 58, 19, 518),
                       dimnames = list(c("Rash +", "Rash -"),
                                       c("Allopurinol +", "Allopurinol -")),
                       nrow = 2, byrow = TRUE)
```

```{r miettinen1-conftab, echo=FALSE}
knitr::kable(rash_conf, caption = "Outcome by confounders")
knitr::kable(rash_males, caption = "Among males")
knitr::kable(rash_females, caption = "Among females")
```

And then get the bias parameters:

```{r miettinen1_parms}
(RR_no <- (36/(36+645))/(58/(58+518))) # RR between confounder and outcome among non-exposed
(p1 <- (5+33)/(15+52)) # prevalence of confounder among exposed
(p2 <- (36+645)/(94+1163)) # prevalence of confounder among unexposed
```

With the following results:

```{r miettinen1_res}
rash %>% confounders(., type = "RR", bias_parms = c(RR_no, p1, p2))
```

For the second example, data are from a case-control study of the relationship
of oral contraceptive use to venous thrombosis with age as a confounding factor.

```{r miettinen2}
thromb <- matrix(c(12, 30, 53, 347),
                 dimnames = list(c("Thrombosis +", "Thrombosis -"),
                                 c("OC +", "OC -")),
                 nrow = 2, byrow = TRUE)
```

```{r miettinen2-tab, echo=FALSE}
knitr::kable(thromb)
```

```{r miettinen2-conf}
## Outcome by confounders
thromb_conf <- matrix(c(18, 12, 158, 189),
                      dimnames = list(c("Thrombosis +", "Thrombosis -"),
                                      c("20-29 years old", "30-39 years old")),
                      nrow = 2, byrow = TRUE)
## By confounders: among 20-29 years old
thromb_younger <- matrix(c(10, 18, 39, 158),
                       dimnames = list(c("Thrombosis +", "Thrombosis -"),
                                       c("OC +", "OC -")),
                       nrow = 2, byrow = TRUE)
## By confounders: among 30-39 years old
thromb_older <- matrix(c(2, 12, 14, 189),
                     dimnames = list(c("Thrombosis +", "Thrombosis -"),
                                     c("OC +", "OC -")),
                     nrow = 2, byrow = TRUE)
```

```{r miettinen2-conftab, echo=FALSE}
knitr::kable(thromb_conf, caption = "Outcome by confounders")
knitr::kable(thromb_younger, caption = "Among 20-29 years old")
knitr::kable(thromb_older, caption = "Among 30-39 years old")
```

And then the bias parameters:

```{r miettinen2_parms}
(OR_no <- (18/12)/(158/189)) # OR between confounder and outcome among non-exposed
(p1 <- (10+39)/(12+53)) # prevalence of confounder among exposed
(p2 <- (18+158)/(30+347)) # prevalence of confounder among unexposed
```

With the following results:

```{r miettinen2_res}
thromb %>% confounders(., type = "OR", bias_parms = c(OR_no, p1, p2))
```

## Bootstrapping

Selection and misclassification bias models can be bootstrapped in order to get
confidence interval on the adjusted association parameter.
We can use the ICU dataset from Hosmer and
Lemeshow [Applied Logistic
Regression](https://ca.wiley.com/WileyCDA/WileyTitle/productCd-0470582472.html)
textbook as an example.
The replicates that give negative cells in the adjusted 2X2 table are silently
ignored and the number of effective bootstrap replicates is provided in the
output.
Ten thousands bootstrap samples are a good number for testing everything runs
smoothly.
But again, don't be afraid of running more, like 100,000 bootstrap samples.

```{r, boot}
library(aplore3)  # to get ICU data
data(icu)

## First run the model
misclass_eval <- misclassification(icu$sta, icu$inf,
                                   type = "exposure",
                                   bias_parms = c(.75, .85, .9, .95))
misclass_eval

## Then bootstrap it
set.seed(123)
misclass_boot <- boot.bias(misclass_eval, R = 10000)
misclass_boot
```

Bootstrap replicates can also be plotted, with the confidence interval shown as
dotted lines.

```{r, boot_fig, fig.cap = "Bootstrap replicates and confidence interval.", warning=F}
plot(misclass_boot, "rr")
```
